Trade-offs among plant floral and reproductive traits determine interactions with floral visitors

### 2021-08-11 Draft/Manuscript.Rmd
manuscript

---
title: "Plant reproductive strategies structure plant-pollinator networks"
author:
  - Jose B. Lanuza^1,2^*
  - Romina Rader^1^
  - Jamie Stavert^3^
  - Liam K. Kendall^4^
  - Manu E. Saunders^1^
  - Ignasi Bartomeus^2^

@@

output:
  bookdown::pdf_document2:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    df_print: kable
    keep_tex: yes
    number_sections: no
    toc: no
    includes:
      in_header: header.tex

documentclass: article
# classoption: twocolumn, landscape, etc
papersize: a4
pagestyle: plain  # plain, empty, headings
geometry: margin=1in
linestretch: 2
fontfamily: mathpazo
fontsize: 11pt

@@

header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
---

@@

<!--# Write Abstract and Keywords below -->

\vspace{1cm}
\hrule

#Abstract

Increasing availability of large datasets have improved our understanding of organismal life history and pollination biology. However, most attention has focused on vegetative traits when delimiting plant life strategies and overlooked key floral and reproductive traits. In addition, plant-pollinator research rarely considers the reproductive biology of the species and has overemphasized the study of highly specialized interactions. Greater knowledge of the plant reproductive trade-offs between these traits can shed light on the macroecological patterns and help our understanding of life strategies.

Here, we compiled 20 reproductive traits for 1,506 plant species from 28 plant-pollinator network studies and 18 different countries that constitute a total of 64 unique networks and 8 meta-webs. We explore the different trade-offs of floral, reproductive, and vegetative traits of the main quantitative traits in the trait space. Then, we evaluate whether the frequency of pollinator visits within networks change across the primary axes of trait variation and pollinator functional groups. Finally, we investigate if the functional role of the plant species (interaction frequency, normalized degree and specialization) in the network is associated with the primary axes of trait variation and the different plant traits.

We found that 51.8% of the trait variation was explained by two independent axes that determined plant form and function. The first axis showed the flower number versus flower size compromise (PC1, 26.72%) where species were divided in two opposed strategies with high investment in flower number and plant height on one extreme or high investment in flower size, style length and larger number of ovules on the other. The second axis showed the pollinator dependence trade-off (PC2, 25.08%) between species with high pollinator dependence with greater plant height and higher investment in floral traits on one end and low pollinator dependence with smaller height and lower investment in floral traits on the other. An additional principal component analysis with the subset of species with quantitative data of pollen and nectar showed that nectar quantity was positively associated with flower size, style length and ovule number (PC1, 24.70%) and pollen quantity was positively correlated with flower number and plant height and negatively associated with autonomous selfing (PC2, 20.59%). The main axes of trait variation with the full set of species did not explain the overall visitation patterns but showed marked trends for the different functional groups. For instance, all pollinator functional groups had higher frequency of visits on plant species with high pollinator dependence (PC2); and bees (Anthophila), the most abundant functional group, had higher frequency of visits on species with high number of flowers and tall plant height (PC1); and, on species with longer styles and higher levels of autonomous selfing (PC3). Overall, plant species with small size, low nectar concentration and short styles showed lower visitation rates than plant species with the opposed traits; species with small amounts of floral rewards (pollen and nectar), low number of flowers and small flowers had in general lower number of visiting species (normalized degree); and species under 1.3 m, with high rewards of pollen and nectar showed the highest levels of specialization. This work highlights the need to consider essential plant reproductive and floral traits to improve understanding of plant life strategies and plant-pollinator interactions at large scales.

\vspace{3mm}
\hrule
\vspace{5mm}

*Keywords*: plant-pollinator networks | floral traits | life history strategy | functional traits

\bleft
\newpage

@@

```{r knitcitations, cache = FALSE}
library(knitcitations)
cleanbib()   
cite_options(citation_format = "pandoc")
```

<!--# Start writing your paper below -->

# INTRODUCTION

>>>>

---
title: "Trade-offs among plant floral and reproductive traits determine interactions with floral visitors"

author:
  - Jose B. Lanuza^1,2^*
  - Romina Rader^1^
  - Jamie Stavert^3^
  - Liam K. Kendall^4^
  - Manu E. Saunders^1^
  - Ignasi Bartomeus^2^

@@

output:
  bookdown::pdf_document2:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    df_print: kable
    keep_tex: yes
    number_sections: no
    extra_dependencies: ["float"]
    toc: no
    fig_crop: no
    includes:
      in_header: header.tex

documentclass: article
# classoption: twocolumn, landscape, etc
papersize: a4
pagestyle: plain  # plain, empty, headings
geometry: margin=1in
linestretch: 2
fontfamily: mathpazo
fontsize: 12pt

@@

header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{float}
- \floatplacement{figure}{H}
- \usepackage[skip=5pt]{caption}
---

@@

Plant life strategies are often delimited by vegetative and physiological traits but little is known about how floral and reproductive traits drive these strategies, and in turn shape plant interactions with floral visitors. Here, we compiled 13 floral, 4 reproductive and 3 vegetative traits for 1,506 plant species from 28 plant-pollinator network studies across 18 different countries. We investigated the associations among these traits, pollinator visitation and the functional role of plant species within the networks (interaction frequency, normalized degree and specialization). We found that 51.8% of trait variation was explained by two independent axes that encompassed plant form and function. Specifically, the first axis indicated the presence of a trade-off between flower number and flower size (PC1, 26.72%). The second axis indicated a trade-off for the level of pollinator dependency (PC2, 25.08%). Although the main axes of trait variation did not fully explain pollinator visitation rates, different plant life strategies were associated with visitation rates and pollinator functional groups. Overall, the main traits that determined plant species’ functional roles were height, nectar concentration, pollen grains per flower, number of ovules, style length, selfing level and flower width. Our results highlight the need to consider plant reproductive and floral traits to improve understanding of plant life strategies and plant-pollinator interactions at broader spatial scales.

\vspace{5mm}
\bleft

@@

```{r knitcitations, cache = FALSE}
library(knitcitations)
cleanbib()   
cite_options(citation_format = "pandoc")

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

There is an astonishing diversity of floral structures and plant reproductive strategies among flowering plants (Barrett, 2002; Schiestl & Johnson, 2013), which have long been of interest to pollination biologists in terms of their relevance to plant-pollinator interactions. However, most studies that have explored reproductive (e.g., mating and compatibility systems) and floral trait (e.g., flower size or nectar provision) variation have concentrated on the individual or community level and thus, broader macroecological patterns remain poorly investigated (Carvalheiro et al., 2014; Baude et al., 2016; Munoz et al., 2016; Moeller et al., 2017; Grossenbacher et al., 2017). Indeed, studies depicting species’ life history strategies generally focus on vegetative traits and rarely consider reproductive traits (Salguero-Gómez et al., 2016; Rüger et al., 2018). As a consequence, a unified framework that explores the compromises among floral traits and their relevance to plant life strategies is currently lacking (Roddy et al., 2021). At the same time, there is growing interest in the determinants of plant-pollinator interactions via trait-based approaches (Fenster et al., 2004) and trait-matching analyses (Bartomeus et al., 2016). However, floral traits have been overlooked beyond highly specialised plant-pollinator systems (Dellinger, 2020; Roddy et al., 2021) and the role of plant reproductive biology remains little explored in plant-pollinator interactions (but see Tur et al., 2013 and Devaux et al., 2014).

With the recent availability of large trait databases, plant ecological strategies are increasingly being examined (e.g., TRY, Kattge et al., 2011 and COMPADRE, Salguero‐Gómez et al., 2015), and are facilitating the identification of global patterns and constraints of plant form and function (Salguero‐Gómez et al., 2015; Díaz et al., 2016; Carmona et al., 2021). However, the main focus has been on vegetative traits such as leaf (Wright et al., 2004) or wood (Chave et al., 2009) trade-offs with little or no attention given to reproductive and floral traits (E‐Vojtkó et al., 2020), also critical to plant form and function. For instance, short lived versus perennial species tend to have low versus high levels of outcrossing, respectively (Barrett, 2003; Moeller et al., 2017). Further, outcrossing levels are positively correlated with flower size (Goodwillie et al., 2010). In addition, the presence of costly rewards (e.g., pollen or nectar) and showy flowers or floral displays can only be understood through consideration of plant species’ reliance upon animal pollination (pollinator dependence) and their role in attracting pollinators (Ollerton et al., 2011). Hence, exploring plant life strategies with reproductive and floral trade-offs, in conjunction with their pollinator dependence, is necessary for a balanced understanding of plant economics.

Several studies have identified links between plant traits and plant-pollinator network properties (Lázaro et al., 2008; Bartomeus, 2013; Carvalheiro et al., 2014). Moreover, plant traits can also define species' network roles (e.g., specialists vs generalists). For example, species that occupy trait space extremes are more likely to exhibit higher specialization and be more reliant on trait-matching (Junker et al., 2013; Coux et al., 2016). This morphological matching between plant and floral visitors can determine plant-pollinator interactions, and thus shape their interaction network structure (Stang et al., 2009; Ibanez, 2012). Despite the increasing knowledge of the relevance of traits on the species network roles, little is known about how plant reproductive and floral traits determine plant species’ network roles at a macroecological scale.

Here, we explore the potential trade-offs among plant floral and reproductive traits and how these influence the structure of plant-pollinator networks. First, we identify the major axes of floral and reproductive trait variation and trade-offs that determine plant form and function. Second, we investigate how plant species’ position in trait-space influences interaction strength with different guilds of floral visitors. Finally, we investigate how the main axes of trait variation and individual traits influence plant species roles within networks using complementary interaction network metrics (i.e., interaction strength, normalized degree and specialization).

**Plant strategies**

The phylogenetically informed principal component analysis (pPCA) captured by the first two and three axes 51.8% and 70.97% of trait variation, respectively (Figure 1 and Figure S5) and had a phylogenetic correlation (**lambda**) of 0.76. The first principal component (PC1) represented 26.72% of the trait variation and indicated a trade-off between flower number and flower size. We refer to this axis as the ‘flower number - flower size trade-off as already described in previous studies (Sargent et al., 2007; Kettle et al., 2011). Hence, one end of the spectrum comprised species with high investment in flower number and plant height but small flower size, short style length and low ovule number. The other end of this spectrum comprised species that were short in height and invested in large flowers, long styles, many ovules, but few flowers. The main contributing traits to PC1 were plant height, flower number, ovule number and flower size (loadings > |0.5|; Table S3) but style length also contributed moderately on PC1 (loading = -0.33). The second principal component (PC2) represented 25.05% of the trait variation and indicated a trade-off between low and high pollinator dependence. We refer to this axis as the ‘pollinator dependence trade-off’. The main driver of trait variation on PC2 was autonomous selfing (loading = 0.85) but the other traits (except ovule number) also made moderate contributions (loadings from 0.27 to 0.4; Table S3). We found that high pollinator dependence was associated with larger and higher numbers of flowers, greater plant height and longer styles. In contrast, species with high levels of autonomous selfing tended to have fewer and smaller flowers, had shorter styles and were shorter in height. Further, PC3 explained a considerable amount of trait variability (19.17%) and the main contributors to this axis were style length (loading = -0.66) and the degree of autonomous selfing (loading = -0.51). The remaining traits, apart from ovule number, were moderately correlated to changes on PC3 (loadings from -0.23 to -0.46; Table S3). Thus, because style length was correlated with all traits on PC3 and was the main driver of variation, we refer to this axis as the ‘style length trade-off’. Further, the pPCA with the subset of species that had nectar and pollen quantity data showed that nectar quantity (microlitres of nectar per flower) was positively associated with flower size, style length and ovule number (PC1, 23.40%); and pollen quantity (pollen grains per flower) was positively correlated with flower number and plant height and negatively associated with autonomous selfing (PC2, 21.67%; Figure S6). This pPCA explained similar variance with the first two principal components (45.07%) and similar associations of traits despite some variability in the loadings (Table S4).

```{r, echo=FALSE, message=FALSE, cache=FALSE, results='hide', warning=FALSE, fig.height=10, fig.width=10,fig.pos = "!H",fig.cap= "Figure 1. Life history strategies for 1,236 plant species from 28 plant-pollinator networks studies across the first two axes of trait variation from a phylogenetically informed principal component analysis (pPCA). The solid arrows indicate the direction of the different quantitative traits (flower number, plant height, style length, flower size, ovule number and level of autonomous selfing) in the 2-d plane. The length of the arrows indicate the weight of the variables on each principal component and the labelled icons at their end represent the extreme form of the trait continuum. The dashed lines show the opposed direction of trait variation and the non-labelled icons at their end illustrate the opposing extreme of the continuum."}

setwd("~/R_Projects/Reproductive Traits")

#Load libraries
library(viridis) #COLOUR GGPLOT
library(ggplot2)
library(broman) #crayon colours
library(cowplot)
library(magick)

# Theme for publication
theme_ms <- function(base_size=12, base_family="Helvetica") {
  (theme_bw(base_size = base_size, base_family = base_family)+
      theme(text=element_text(color="black"),
            axis.title=element_text( size = rel(1.75)),
            axis.text=element_text(size = rel(1.75), color = "black"),
            legend.title=element_text(face="bold"),
            legend.text=element_text(),
            legend.background=element_rect(fill="transparent"),
            legend.key.size = unit(1.2, 'lines'),
            panel.border=element_rect(color="black",size=1),
            panel.grid.minor.x =element_blank(),
            panel.grid.minor.y= element_blank(),
             panel.grid.major= element_blank()
      ))
}

  plot <- plot + xlab(paste("Flower number - flower size trade-off ", "(",(percentage[1]),"%",")", sep = "")) #XLAB
  plot <- plot + ylab(paste("Pollinator dependence trade-off ", "(",(percentage[2]),"%",")", sep = "")) #YLAB

  plot <- plot + theme_bw() + annotation_raster(tree_1, xmin = -4.1, xmax = -2.85,ymin = 2.05, ymax = 4.05)+
     annotation_raster(style_l_1, xmin = 2.25, xmax = 3.05,ymin = 2.55, ymax = 3.7) +
    annotation_raster(style_s_1, xmin = -2.6, xmax = -1.9,ymin = -2.65, ymax = -1.7) +
    annotation_raster(ovule_h_1, xmin = 3.5, xmax = 4.2,ymin = -0.45, ymax = 0.50) +
    annotation_raster(ovule_l_1, xmin = -4.2, xmax = -3.5,ymin = -0.45, ymax = 0.5) +
    annotation_raster(selfing_n_1, xmin = -0.75, xmax = 1.2,ymin = 2, ymax = 4.7) +
    annotation_raster(selfing_h_1, xmin = -0.30, xmax = 0.15,ymin = -3.9, ymax = -3.1) +
    annotation_raster(single_flower_1, xmin = 3.3, xmax = 4.1,ymin = -1.7, ymax = -0.7) +
    annotation_raster(many_flowers_1, xmin = -4.20, xmax = -3.40,ymin = 0.8, ymax = 1.45) +
    annotation_raster(large_flower_1, xmin = 3.15, xmax = 4.15,ymin = 1.3, ymax = 2.55) +
    annotation_raster(flower_small_1, xmin = -3.50, xmax = -2.55,ymin = -2, ymax = -1.50) +
    annotation_raster(herb_1, xmin = 2.6, xmax = 3,ymin = -2.8, ymax = -2)

      #ADD LABELS
  rownames(PC$L) <- c("High selfing", "Many flowers", "Large flowers", "Long styles", "Many ovules", "Tall plants" )

  PCAloadings <- data.frame(Variables = rownames(PC$L), PC$L)
 plot <- plot + annotate("text", x = -(PCAloadings$PC1*c(4.7,5.05,5.6,7.85,7.3,6.15)), y = -(PCAloadings$PC2*c(4.693,2.5,3.1,6.6,5,5.5)+c(0,0,0,0,0.4,0)),
                         label = PCAloadings$Variables, color="black",fontface =2,size=4)

 plot <- plot + annotate("text", x = (PCAloadings$PC1*c(4.6,4.95,4.5,6.6,7.3,5.15)), y = (PCAloadings$PC2*c(3.83,6.4,5.7,7.4,-1,7.45)+c(0,0,0,0,-0.41,0)),
                         label =c("Low selfing", "Few flowers", "Small flowers", "Short styles", "Few ovules", "Short plants" ), color="black",fontface =2,size=4)

We found that most categorical traits were statistically associated with the first two axes of trait variation (Figure 2 and Table S2). Flower symmetry, which was only associated with PC2 (Sum of squares = 8.51, F-value = 14.72, p < 0.01 ), and nectar provision, which was independent of PC1 and PC2 (PC1: Sum of squares = 0.37, F-value = 0.29 , p = 0.59; PC2:  Sum of squares = 0.83, F-value = 1.43, p = 0.23) showed lack of statistical association. In addition, we found statistical differences (Tukey test) between the different levels of categorical traits in the trait space (Figure S7). Regarding self compatibility, we found larger differences on PC2 (i.e., species with unisexual flowers that were self incompatibile were statistically differentiated from species with partial or full self compatibility; Figure S7 A and B). Life forms differed statistically across both axes of trait variation and followed a gradient of larger life forms (trees and shrubs) with higher pollinator dependence to smaller ones (herbs) with lower pollinator dependence (Figure S7 C and D). Consequently, lifespan also followed this gradient but perennial and short lived species only differed statistically on PC2 (Figure S7 E and F). Species with unisexual flowers (monoecious and dioecious) were clustered on both extremes of the first two principal components and had the highest pollinator dependence and highest number of flowers (Figure S7 G and H). Moreover, we found that the campanulate and capitulum flower shapes were differentiated from tube, papilionaceous, open and brush shapes in the trait space. The former morphologies had larger flowers and greater pollinator dependence, while the latter had higher flower number and greater autonomous selfing (Figure S7 I and J). Regarding flower symmetry, zygomorphic flowers were associated with lower levels of pollinator dependence, whereas actinomorphic flowers had higher levels of pollinator dependence (Figure S7 K and L).

\blandscape

```{r, echo=FALSE, message=FALSE, cache=FALSE, results='hide', warning=FALSE, fig.height=13, fig.width=27, out.extra='',fig.cap="Location of the different qualitative traits that showed statistical association with the first two axes of trait variation. These different traits included: compatibility system (a), life form (b), lifespan (c), breeding system (d), flower shape (e) and flower symmetry (f)."}

#LOAD LIBRARIES 
#library(phytools)
#library(ape) #for phylogenetic distance
#library(dplyr) #data processing
#library(rtrees) #for phylogenetic distancelibrary(MASS)
#library(reshape2)
library(viridis) #COLOUR GGPLOT
#library(MASS)
library(ggplot2)
library(broman) #crayon colours
#library(tidyverse)
library(patchwork) #panel of plots
#library(brms)
#library(cmdstanr)

setwd("~/R_Projects/Reproductive Traits")

p1 <- readRDS("Main_Figures/figure_2_1.rds")
p2 <- readRDS("Main_Figures/figure_2_2.rds")
p3 <- readRDS("Main_Figures/figure_2_3.rds")
pp1 <- readRDS("Main_Figures/figure_2_4.rds")
pp2 <- readRDS("Main_Figures/figure_2_5.rds")
pp3 <- readRDS("Main_Figures/figure_2_6.rds")

\elandscape

**Phylogenetic signal of traits**

###OLD
